---
title: "Figure S8: Similarity between BE and all other tissues"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here, we perform additional analyses on the similarity between BE and all other tissues.

# Read in the data

```{r libraries, message=FALSE}
library(scran)
library(scater)
library(DropletUtils)
library(openxlsx)
library(Rtsne)
library(pheatmap)
library(viridis)
library(ape)
library(batchelor)
source("../../Analysis/Functions/auxiliary.R")

sce <- readRDS("../../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")

# Exclude contaminating cells
sce <- sce[,sce$include]
```

# Plot circular dendrogram of all clusters

To estimate the similarity between all cell types, we compute an averaged expression profile for each cell-type and perform hierarchical cluster analysis.
Here, we calculate this on the non-batch corrected counts and thereby incorporating the inter-patient variation.
We will first plot the dendogram across all genes, removing lowly expressed genes

```{r}
# Remove non-expressed genes
sce <- sce[Matrix::rowMeans(counts(sce)) > 0,]

# Normalize across all cells
# We provide the cell-types as clusters
sce <- computeSumFactors(sce, clusters=paste(sce$Tissue, sce$cell_type, sep = "_"))
sce <- normalize(sce, return_log = FALSE)

clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")
mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(sce))
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- Matrix::rowMeans(normcounts(sce)[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate spearman correlation between cell types
dend <- hclust(as.dist(sqrt(1 - cor(log2(mat + 1), method = "spearman"))/2), method = "ward.D2")

dev.off()
pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_all_genes.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```

Next we will use the same genes as the ones for batch correction.

```{r}
clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")

# Find highly variable genes
# Load sce again to keep internal normalization intact
sce.new <- readRDS("../../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")
sce.new <- sce.new[,sce.new$include]
sce.list <- split.sce(sce.new, unique(sce.new$Sample), colData.name = "Sample")

HVG.genes <- lapply(sce.list, function(n){
  HVG <- trendVar(n, use.spikes = FALSE)
  decomposeVar(n, HVG)
})
  
HVG.df <- do.call("combineVar", HVG.genes)
HVG.df <- HVG.df[order(HVG.df$bio, decreasing = TRUE),]
genes <- rownames(HVG.df)[1:1000]

mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = 1000)
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- Matrix::rowMeans(normcounts(sce)[genes,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate spearman correlation between cell types
dend <- hclust(as.dist(sqrt(1 - cor(log2(mat + 1), method = "spearman"))/2), method = "ward.D2")

dev.off()
pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_HVG.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```

# Multi-batch scaling

We next scale the batches as a way to remove batch effects. 

```{r}
# Read in the data
sce <- readRDS("../../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")

# Exclude contaminating cells
sce <- sce[,sce$include]

# Split the object into batches
sce.list <- split.sce(sce, unique(sce$Sample), colData.name = "Sample")

# Scale batches
library(batchelor)
func <- paste0("batchelor::multiBatchNorm(", 
                     paste0("sce.list[[", 1:length(sce.list), "]]", collapse=", "), 
                     ")")
sce.scaled <- eval( parse(text=func) )

# Merge these datasets
sce <- do.call("cbind", sce.scaled)

# Remove non-expressed genes
sce <- sce[Matrix::rowMeans(counts(sce)) > 0,]

# Normalize across all cells
# We provide the cell-types as clusters
sce <- computeSumFactors(sce, clusters=paste(sce$Tissue, sce$cell_type, sep = "_"))
sce <- normalize(sce, return_log = FALSE)
```

We next repeat the analysis as above.
First, we calculate a consensus transcriptome for each cell-type based on all genes.

```{r}
clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")
mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(sce))
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- Matrix::rowMeans(normcounts(sce)[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate spearman correlation between cell types
dend <- hclust(as.dist(sqrt(1 - cor(log2(mat + 1), method = "spearman"))/2), method = "ward.D2")

dev.off()
pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_all_genes.batch_scaled.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```

Now with the HVGs.

```{r}
clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")

# Find highly variable genes
# Load sce again to keep internal normalization intact
sce.new <- readRDS("../../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")
sce.new <- sce.new[,sce.new$include]
sce.list <- split.sce(sce.new, unique(sce.new$Sample), colData.name = "Sample")

HVG.genes <- lapply(sce.list, function(n){
  HVG <- trendVar(n, use.spikes = FALSE)
  decomposeVar(n, HVG)
})
  
HVG.df <- do.call("combineVar", HVG.genes)
HVG.df <- HVG.df[order(HVG.df$bio, decreasing = TRUE),]
genes <- rownames(HVG.df)[1:1000]

mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = 1000)
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- Matrix::rowMeans(normcounts(sce)[genes,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate spearman correlation between cell types
dend <- hclust(as.dist(sqrt(1 - cor(log2(mat + 1), method = "spearman"))/2), method = "ward.D2")

dev.off()
pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_HVG.batch_scaled.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```

# Different number of HVG

Here, I will use different sets of genes to show that the batch correction is independent from the chosen gene set.

```{r batch-correction}
detach("package:batchelor", unload=TRUE)
all_sce <- readRDS("../../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")
all_sce <- all_sce[,all_sce$include]
sce.list <- split.sce(sce = all_sce, groups = unique(all_sce$Sample), colData.name = "Sample")

# Combine sce objects
all_sce <- do.call("cbind", sce.list)
  
# Batch correction
corrected_1 <- batch.correction(sce.list, number.HVG = 500)
corrected_2 <- batch.correction(sce.list, number.HVG = 2000)

# Compute tsne on corrected counts
set.seed(1234)
tsne_1 <- Rtsne(t(corrected_1), pca = FALSE)
tsne_2 <- Rtsne(t(corrected_2), pca = FALSE)

# dendogram analysis
clusters <- paste(colData(all_sce)$Tissue, colData(all_sce)$Tissue_cluster, sep = "_")
mat_1 <- mat_2 <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(corrected_1))
colnames(mat_1) <- colnames(mat_2) <- unique(clusters)

for(i in unique(clusters)){
  mat_1[,i] <- rowMeans(corrected_1[,clusters == i])
  mat_2[,i] <- rowMeans(corrected_2[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat_1)){
  colnames(mat_1)[i] <- unique(all_sce$cell_type[all_sce$Tissue == sub("_[0-9]*$", "", colnames(mat_1)[i]) &
                                 all_sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat_1)[i])])
  colnames(mat_2)[i] <- unique(all_sce$cell_type[all_sce$Tissue == sub("_[0-9]*$", "", colnames(mat_2)[i]) &
                                 all_sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat_2)[i])])
}

# Calculate euclidean distance on corrected counts
dend_1 <- hclust(dist(t(mat_1), method = "euclidean"), method = "ward.D2")
dend_2 <- hclust(dist(t(mat_2), method = "euclidean"), method = "ward.D2")
dev.off()
pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_500HVG.pdf", width = 10, height = 10)
plot(as.phylo(dend_1), type = "fan", 
     tip.color = metadata(all_sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()

pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_2000HVG.pdf", width = 10, height = 10)
plot(as.phylo(dend_2), type = "fan", 
     tip.color = metadata(all_sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```

# Patient correction

Next, I will only correct across the different patients.

```{r}
all_sce <- readRDS("../../../../Dropbox (Personal)//Oesophagus_single_cell/All_corrected_sce.rds")
all_sce <- all_sce[,all_sce$include]
sce.list <- split.sce(sce = all_sce, groups = unique(all_sce$Patient), colData.name = "Patient")

# Correctly normalize for each patient
sce.list <- lapply(sce.list, function(n){
  cur_cluster <- quickCluster(n, method = "igraph", irlba.args = c("work" = 100))
  n <- computeSumFactors(n, clusters=cur_cluster)
  n <- normalize(n, return_log = TRUE)
  n
})

# Order list
sce.list <- sce.list[c("Patient2", "Patient3", "Patient7", "Patient9", "Patient8", "Patient10", "Patient12", "Patient11", "Patient13")]

# Combine sce objects
all_sce <- do.call("cbind", sce.list)

# Perform batch correction
corrected <- batch.correction(sce.list)

set.seed(1234)
tsne <- Rtsne(t(corrected), pca = FALSE, perxplexity = 100)

# dendogram analysis
clusters <- paste(colData(all_sce)$Tissue, colData(all_sce)$Tissue_cluster, sep = "_")
mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(corrected))
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- rowMeans(corrected[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(all_sce$cell_type[all_sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 all_sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate euclidean distance on corrected counts
dend <- hclust(dist(t(mat), method = "euclidean"), method = "ward.D2")
dev.off()
pdf("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S8/Tree_patient_correction.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(all_sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```
