---
title: "Figure S6: Differential expression analysis between tissues"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script focuses on detecting the genes that are differentially expressed between BE and SCJ, GC and GOJ, NE and SCJ and NE and GOJ

# Read in the data

```{r libraries, message=FALSE}
library(scran)
library(scater)
library(DropletUtils)
library(openxlsx)
library(Rtsne)
library(pheatmap)
library(RColorBrewer)
library(edgeR)
library(viridis)
source("../../Analysis/Functions/auxiliary.R")
sce <- readRDS("../../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")
```

# Visualize GC, GOJ, BE and SCJ samples

Here we perform differential expression (DE) analysis between the epithelial tissues of the GC, GOJ, SCJ, BE and NE samples.

```{r}
# Select tissues to test
cur_sce <- sce[,sce$include & sce$Tissue %in% c("NE", "GC", "BE", "GOJ", "SCJ")]

# Perform batch correction
sce.list <- split.sce(cur_sce, unique(cur_sce$Sample), colData.name = "Sample")
corrected <- batch.correction(sce.list)
cur_sce <- do.call("cbind", sce.list)

# Compute new tSNE
set.seed(11111)
tsne <- Rtsne(t(corrected), pca = FALSE, perplexity = 100)
reducedDims(cur_sce)$TSNE <- tsne$Y

# Visualize tSNEs
# Tissue
p.all.cells <- ggplot(data.frame(tsne1 = reducedDims(cur_sce)$TSNE[,1],
                  tsne2 = reducedDims(cur_sce)$TSNE[,2],
                  tissue = colData(cur_sce)$Tissue)) + 
  geom_point(aes(tsne1, tsne2), colour = "black", size = 1) + 
  geom_point(aes(tsne1, tsne2, colour = tissue), size = 0.5) + 
  scale_color_manual(values = metadata(cur_sce)$colour_vector) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "grey"))
ggsave("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S6/All_cells.pdf", p.all.cells, width = 7, height = 5)

# Tissue typse
p.all.cells.tissue_type <- ggplot(data.frame(tsne1 = reducedDims(cur_sce)$TSNE[,1],
                  tsne2 = reducedDims(cur_sce)$TSNE[,2],
                  tissue = colData(cur_sce)$tissue_type)) +
  geom_point(aes(tsne1, tsne2), colour = "black", size = 1) + 
  geom_point(aes(tsne1, tsne2, colour = tissue), size = 0.5) + 
  scale_color_manual(values = c("dark blue", "dark red")) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "grey"))
ggsave("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S6/Tissue_type.pdf", p.all.cells.tissue_type, width = 7, height = 5)

# Cell type
colour_vector <- vector(length = length(unique(cur_sce$cell_type)))
names(colour_vector) <- unique(cur_sce$cell_type)
colour_vector["Unknown"] <- "grey20"
colour_vector["Superficial"] <- colorRampPalette(c("white", "dark red"))(10)[10]
colour_vector["Basal"] <- colorRampPalette(c("white", "dark red"))(10)[4]
colour_vector["Intermediate"] <- colorRampPalette(c("white", "dark red"))(10)[7]
colour_vector["Dividing"] <- colorRampPalette(c("white", "dark red"))(10)[1]
colour_vector["Foveolar_differentiated"] <- colorRampPalette(c("white", "#4DBBD5FF"))(10)[10]
colour_vector["Endocrine"] <- colorRampPalette(c("white", "dark blue"))(10)[10]
colour_vector["Undifferentiated"] <- colorRampPalette(c("white", "#4DBBD5FF"))(10)[3]
colour_vector["Foveolar_Intermediate"] <- colorRampPalette(c("white", "#4DBBD5FF"))(10)[6]
colour_vector["Chief"] <- colorRampPalette(c("white", "dark blue"))(10)[6]
colour_vector["Parietal"] <- colorRampPalette(c("white", "dark blue"))(10)[3]
colour_vector["Foveolar_Intermediate.goblet"] <- "black"
colour_vector["Endocrine.Undifferentiated"] <- "grey60"
colour_vector["Goblet"] <- colorRampPalette(c("white", "seagreen4"))(10)[10]

p.all.cells.cell_type <- ggplot(data.frame(tsne1 = reducedDims(cur_sce)$TSNE[,1],
                  tsne2 = reducedDims(cur_sce)$TSNE[,2],
                  tissue = colData(cur_sce)$cell_type)) +
  geom_point(aes(tsne1, tsne2), colour = "black", size = 1) + 
  geom_point(aes(tsne1, tsne2, colour = tissue), size = 0.5) + 
  scale_color_manual(values = colour_vector) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "grey"))
ggsave("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S6/Cell_type.pdf", p.all.cells.cell_type, width = 7, height = 5)

# Visualize patient
# Colouring for patient
colour_vector <- vector(length = length(unique(cur_sce$Patient)))
names(colour_vector) <- unique(cur_sce$Patient)
colour_vector <- brewer.pal(7, "Accent")

p.all.cells.patient <- ggplot(data.frame(tsne1 = reducedDims(cur_sce)$TSNE[,1],
                  tsne2 = reducedDims(cur_sce)$TSNE[,2],
                  tissue = colData(cur_sce)$Patient)) +
  geom_point(aes(tsne1, tsne2), colour = "black", size = 1) + 
  geom_point(aes(tsne1, tsne2, colour = tissue), size = 0.5) + 
  scale_color_manual(values = colour_vector) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "grey"))
ggsave("../../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S6/Patient.pdf", p.all.cells.patient, width = 7, height = 5)
```

# Differential expression analysis

Here we berform differential expression analysis between NE and squamous cells of GOJ, GC and columnar cells of GC, NE and squamous cells of SCJ, BE and columnar cells of SCJ.
For this, we will sum the raw counts for each gene across all cells per tissue and per patient. 
We then use edgeR on these pseudo-bulk samples to test for changes in mean expression.

```{r}
# Define conditions to test
# First two slots are tissues and last slot is cell-types to compare
cond <- list(c("NE", "GOJ", "Basal-Intermediate-Superficial-Dividing"),
             c("GC", "GOJ", "Foveolar_Intermediate-Foveolar_differentiated-Endocrine-Undifferentiated-Chief"),
             c("BE", "SCJ", "Foveolar_differentiated-Foveolar_Intermediate-Undifferentiated-Dividing"),
             c("NE", "SCJ", "Basal-Intermediate-Superficial-Dividing"))

for(i in 1:length(cond)){
  tissue1 <- cond[[i]][1]
  tissue2 <- cond[[i]][2]
  cell_types <- unlist(strsplit(cond[[i]][3], "-"))
  
  # Select cells
  sce_to_test <- cur_sce[,(cur_sce$Tissue == tissue1 | 
                             cur_sce$Tissue == tissue2) &
                           cur_sce$cell_type %in% cell_types]
  
  # Remove non-expressed genes 
  sce_to_test <- sce_to_test[Matrix::rowSums(counts(sce_to_test)) > 0,]
  
  # Peform DE
  cur_out <- DE.edgeR(sce = sce_to_test,
                      conditions = sce_to_test$Tissue,
                      covariate = sce_to_test$Patient,
                      lfc = 0.5,
                      FDR = 0.1)
  
  # Calculate the percentage of DE genes
  print(nrow(do.call("rbind", cur_out))/nrow(sce_to_test))
}

# NE vs GOJ: 0%
# GC vs GOJ: 0.2%
# BE vs SCJ: 1.5%
# NE vs SCJ: 0.2%
```

```{r}
sessionInfo()
```