---
title: "Figure S4: Computational processing of all data"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, 
        encoding = encoding, output_file = '../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Fig_S2.html') })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script visualizes the filtered data obtained after running the Quality_control, Normalization and Further_processing scripts. 
We will first plot these cells to highlight the patient effect and afterwards perform batch-correction.
Finally, we will mark the high-quality batches and cells.

# Read in the data

```{r libraries, message=FALSE}
library(scran)
library(scater)
library(DropletUtils)
library(batchelor)
library(openxlsx)
library(Rtsne)
source("../../Analysis/Functions/auxiliary.R")
```

# Dimensionality reduction

We will first visualize the uncorrected data.

```{r}
# Read in all datafiles
sce.list <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/All_sce.rds")

# Here we exclude the samples that were just used as test runs
sce.list <- sce.list[!(names(sce.list) %in% c("Patient6_SIGAH9_DT1", "Patient6_SIGAG9_GD3",
                                              "Patient4_SIGAG4_SMG"))]

# Remove all unecessary rowData prior to merging
sce.list <- lapply(sce.list, function(n){
    rowData(n) <- rowData(n)[,1:3]
    n
  })

# Merge datasets
sce.all <- do.call("cbind", sce.list)

# Normalize across all cells
clusters <- quickCluster(sce.all, method = "igraph", irlba.args = c("work" = 100))
sce.all <- computeSumFactors(sce.all, clusters=clusters)
sce.all <- normalize(sce.all, return_log = TRUE)

# Compute the top 1000 shared highly variable genes
HVG.genes <- lapply(sce.list, function(n){
    HVG <- trendVar(n, use.spikes = FALSE)
    decomposeVar(n, HVG)
  })
HVG.df <- do.call("combineVar", HVG.genes)
HVG.df <- HVG.df[order(HVG.df$bio, decreasing = TRUE),]
HVgenes <- rownames(HVG.df)[1:1000]

# Compute tsne
set.seed(123456)
tsne <- Rtsne(t(as.matrix(logcounts(sce.all)[HVgenes,])), perplexity = 100)

# Create annotation and colour vector
annot <- paste(colData(sce.all)$Tissue, colData(sce.all)$Patient, sep = "_")
annot.col <- vector(length = length(unique(annot)))
names(annot.col) <- unique(annot)
annot.col[grepl("NE", names(annot.col))] <- colorRampPalette(c("white", "dark red"))(21)[c(5,7,9,11,13,15,17,19,21)]
annot.col[grepl("GOJ", names(annot.col))] <- colorRampPalette(c("white", "#F39B7FFF"))(21)[c(9,11,13,15,17,19,21)]
annot.col[grepl("SCJ", names(annot.col))] <- colorRampPalette(c("white", "#DC0000FF"))(17)[c(13,15,17)]
annot.col[grepl("BE", names(annot.col))] <- colorRampPalette(c("white", "#00A087FF"))(17)[c(5,9,13,17)]
annot.col[grepl("D2", names(annot.col))] <- colorRampPalette(c("white", "#3C5488FF"))(17)[c(5,9,13,17)]
annot.col[grepl("GC", names(annot.col))] <- colorRampPalette(c("white", "#4DBBD5FF"))(25)[c(5,7,9,11,13,15,17,19,21,23,25)]
annot.col[grepl("SMG", names(annot.col))] <- colorRampPalette(c("white", "#B09C85FF"))(17)[c(5,9,13)]

# Plot tsne
tsne.uncorrected <- ggplot(data.frame(tsne1 = tsne$Y[,1], 
                                      tsne2 = tsne$Y[,2], 
                                      Sample = annot)) + 
  geom_point(aes(tsne1, tsne2, colour = Sample)) + 
  scale_colour_manual(values = annot.col) + 
  theme_minimal()

ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S4/All_data_uncorrected.pdf", tsne.uncorrected, width = 12, height = 7)
```

# Batch correction across tissues

We will now perform batch correction across all samples. 
For this, we will use the samples that contain a mix of different tissues first.

```{r}
# Order sce objects for batch correction
n <- names(sce.list)
sce.list <- sce.list[c(which(grepl("GOJ", n)), which(grepl("SCJ", n)),
                       which(grepl("NE", n)), which(grepl("GC", n)),
                       which(grepl("BE", n)), which(grepl("D2", n)),
                       which(grepl("SMG", n)))]

# Combine sce objects
sce.all <- do.call("cbind", sce.list)
  
# Batch correction
corrected <- batch.correction(sce.list)
  
# Compute tsne on corrected counts
set.seed(123)
tsne <- Rtsne(t(corrected), pca = FALSE, perplexity = 150)

# Visualize batches
annot <- paste(colData(sce.all)$Tissue, colData(sce.all)$Patient, sep = "_")

tsne.corrected <- ggplot(data.frame(tsne1 = tsne$Y[,1], tsne2 = tsne$Y[,2], 
                        Sample = annot)) + 
  geom_point(aes(tsne1, tsne2, colour = Sample)) + scale_colour_manual(values = annot.col) + theme_minimal()

ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S4/All_data_corrected.pdf", tsne.corrected, width = 12, height = 7)
```

# High quality samples

In the next step, we will highlight the high quality samples. 
The exclusion criteria is explained in Suppementary Note 1.

```{r}
# Read in the data of high quality samples
sce.high.quality <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/High_quality_sce.rds")
annot.high.quality <- sapply(names(sce.high.quality), function(n){paste(unlist(strsplit(n, "_"))[c(3,1)], collapse = "_")})

# Grey out the cells that we are not using
annot.col.high.quality <- annot.col
annot.col.high.quality[!(names(annot.col.high.quality) %in% annot.high.quality)] <- "black"

# Dataframe for visualization
df <- data.frame(tsne1 = tsne$Y[,1], tsne2 = tsne$Y[,2], 
                 Sample = annot,
                 include = paste(sce.all$Tissue, sce.all$Patient, sep = "_") %in%
                   annot.high.quality)

tsne.corrected.high.quality <- ggplot() + 
  geom_point(data = df[df$include,],
             aes(tsne1, tsne2, colour = Sample)) +
  geom_point(data = df[!df$include,],
             aes(tsne1, tsne2, colour = Sample), shape = 1, alpha = 0.5) +
  scale_colour_manual(values = annot.col.high.quality) + 
  theme_minimal()

ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S4/All_data_corrected_highQuality.pdf", tsne.corrected.high.quality, width = 12, height = 7)
```

# View clusters that were removed due to possible contaminations

In the final step, we remove groups of cells that present contaminations when sampling the tissues.
For example these are squamous oesophagus cells found in the gastric sample.

```{r}
sce.final <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/All_corrected_sce.rds")

# Create new colour vector
annot <- paste(sce.all$Tissue, sce.all$Patient, sep = "_")
annot[!(annot %in% annot.high.quality)] <- "exclude"
annot[match(paste(sce.final$Sample[!sce.final$include], sce.final$Barcode[!sce.final$include], sep = "_"), paste(sce.all$Sample, sce.all$Barcode, sep = "_"))] <- "exclude"

# Grey out the cells that we are not using
annot.col.high.quality <- c(annot.col, "black")
names(annot.col.high.quality)[length(annot.col.high.quality)] <- "exclude"

# Dataframe for visualization
df <- data.frame(tsne1 = tsne$Y[,1], tsne2 = tsne$Y[,2], 
                 Sample = annot)

tsne.corrected.final <- ggplot() + 
  geom_point(data = df[df$Sample == "exclude",],
             aes(tsne1, tsne2, colour = Sample)) +
  geom_point(data = df[df$Sample != "exclude",],
             aes(tsne1, tsne2, colour = Sample)) +
  scale_colour_manual(values = annot.col.high.quality) + 
  theme_minimal()

ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/S4/All_data_corrected_final.pdf", tsne.corrected.final, width = 12, height = 7)
```