---
title: "Figure S5: Similarity between BE and all other tissues"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, 
        encoding = encoding, output_file = '../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Fig_S4.html') })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here, we perform additional analyses on the similarity between BE and all other tissues.

# Read in the data

```{r libraries, message=FALSE}
library(scran)
library(scater)
library(DropletUtils)
library(openxlsx)
library(Rtsne)
library(pheatmap)
library(viridis)
library(ape)
source("../../Analysis/Functions/auxiliary.R")

sce <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/All_corrected_sce.rds")
```

# Plot circular dendrogram of all clusters

To estimate the similarity between all cell types, we compute an averaged expression profile for each cell-type and perform hierarchical cluster analysis.
Here, we calculate this on the non-batch corrected counts and thereby incorporating the inter-patient variation.
We will first plot the dendogram across all genes, removing lowly expressed genes

```{r}
clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")
mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(sce))
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- Matrix::rowMeans(logcounts(sce)[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Remove lowly expressed genes
mat <- mat[rowMeans(mat) > 0.1,]

# Calculate euclidean distance on corrected counts
dend <- hclust(as.dist(sqrt(1 - cor(mat, method = "spearman"))/2), method = "ward.D2")
pdf("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/Figure_S5/Tree_all_genes.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()

# Plot PCA
# Plot first two components
pca <- prcomp(t(mat))
pca.p <- ggplot(data.frame(Component1 = pca$x[,1],
                  Component2 = pca$x[,2],
                  tissue_type = sub("_[0-9]*$", "", unique(clusters)),
                  cell_type = colnames(mat))) +
  geom_point(aes(Component1, Component2, colour = tissue_type)) +
  scale_colour_manual(values = metadata(sce)$colour_vector) + 
  geom_text(aes(Component1, Component2, label = cell_type))

ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/Figure_S5/PCA_cell_types_all_genes.pdf", pca.p, width = 10, height = 10)
```

Next we will use the same genes as the ones for batch correction.

```{r}
clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")

# Find highly variable genes
sce.list <- split.sce(sce, unique(sce$Sample), colData.name = "Sample")

HVG.genes <- lapply(sce.list, function(n){
  HVG <- trendVar(n, use.spikes = FALSE)
  decomposeVar(n, HVG)
})
  
HVG.df <- do.call("combineVar", HVG.genes)
HVG.df <- HVG.df[order(HVG.df$bio, decreasing = TRUE),]
genes <- rownames(HVG.df)[1:1000]

mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = 1000)
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- Matrix::rowMeans(logcounts(sce)[genes,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate euclidean distance on corrected counts
dend <- hclust(as.dist(sqrt(1 - cor(mat, method = "spearman"))/2), method = "ward.D2")
pdf("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/Figure_S5/Tree_HVG.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()

# Plot PCA
# Plot first two components
pca <- prcomp(t(mat))
pca.p <- ggplot(data.frame(Component1 = pca$x[,1],
                  Component2 = pca$x[,2],
                  tissue_type = sub("_[0-9]*$", "", unique(clusters)),
                  cell_type = colnames(mat))) +
  geom_point(aes(Component1, Component2, colour = tissue_type)) +
  scale_colour_manual(values = metadata(sce)$colour_vector) + 
  geom_text(aes(Component1, Component2, label = cell_type))

ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Figures/Supplementary_figures/Figure_S5/PCA_cell_types_HVG.pdf", pca.p, width = 10, height = 10)

```

