---
title: "Validation of batch correction"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, 
        encoding = encoding, output_file = '../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Reports/Batch_correction_validation.html') })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in the data

```{r libraries, message=FALSE}
library(scran)
library(scater)
library(DropletUtils)
library(openxlsx)
library(Rtsne)
library(ape)
source("../../Analysis/Functions/auxiliary.R")
```

# Batch correction across tissues

Here, I will use different sets of genes to show that the batch correction is independent from the chosen gene set.

```{r batch-correction}
all_sce <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/All_corrected_sce.rds")
all_sce <- all_sce[,all_sce$include]
sce.list <- split.sce(sce = all_sce, groups = unique(all_sce$Sample), colData.name = "Sample")

# Combine sce objects
all_sce <- do.call("cbind", sce.list)
  
# Batch correction
corrected_1 <- batch.correction(sce.list, number.HVG = 500)
corrected_2 <- batch.correction(sce.list, number.HVG = 2000)

# Compute tsne on corrected counts
set.seed(1234)
tsne_1 <- Rtsne(t(corrected_1), pca = FALSE)
tsne_2 <- Rtsne(t(corrected_2), pca = FALSE)

p.tsne.500 <- ggplot(data.frame(tSNE1 = tsne_1$Y[,1],
                  tSNE2 = tsne_1$Y[,2],
                  clusters = as.factor(colData(all_sce)$Tissue))) + 
  geom_point(aes(tSNE1, tSNE2, colour = clusters)) +
  scale_colour_manual(values = metadata(all_sce)$colour_vector)
ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/tsne_500HVG.pdf", 
       p.tsne.500, width = 12, height = 7)

p.tsne.2000 <- ggplot(data.frame(tSNE1 = tsne_2$Y[,1],
                  tSNE2 = tsne_2$Y[,2],
                  clusters = as.factor(colData(all_sce)$Tissue))) + 
  geom_point(aes(tSNE1, tSNE2, colour = clusters)) +
  scale_colour_manual(values = metadata(all_sce)$colour_vector)
ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/tsne_2000HVG.pdf", 
       p.tsne.2000, width = 12, height = 7)

# dendogram analysis
clusters <- paste(colData(all_sce)$Tissue, colData(all_sce)$Tissue_cluster, sep = "_")
mat_1 <- mat_2 <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(corrected_1))
colnames(mat_1) <- colnames(mat_2) <- unique(clusters)

for(i in unique(clusters)){
  mat_1[,i] <- rowMeans(corrected_1[,clusters == i])
  mat_2[,i] <- rowMeans(corrected_2[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat_1)){
  colnames(mat_1)[i] <- unique(all_sce$cell_type[all_sce$Tissue == sub("_[0-9]*$", "", colnames(mat_1)[i]) &
                                 all_sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat_1)[i])])
  colnames(mat_2)[i] <- unique(all_sce$cell_type[all_sce$Tissue == sub("_[0-9]*$", "", colnames(mat_2)[i]) &
                                 all_sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat_2)[i])])
}

# Calculate euclidean distance on corrected counts
dend_1 <- hclust(dist(t(mat_1), method = "euclidean"), method = "ward.D2")
dend_2 <- hclust(dist(t(mat_2), method = "euclidean"), method = "ward.D2")
pdf("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/Tree_500HVG.pdf", width = 10, height = 10)
plot(as.phylo(dend_1), type = "fan", 
     tip.color = metadata(all_sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
pdf("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/Tree_2000HVG.pdf", width = 10, height = 10)
plot(as.phylo(dend_2), type = "fan", 
     tip.color = metadata(all_sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```

Next, I will use scanorama and Seurat to perform the same mapping strategy.
Start with scanorama.

```{r}
library(reticulate)
scanorama <- import('scanorama')

# List of genes
HVG.genes <- lapply(sce.list, function(n){
    HVG <- trendVar(n, use.spikes = FALSE)
    decomposeVar(n, HVG)
})
  
HVG.df <- do.call("combineVar", HVG.genes)
HVG.df <- HVG.df[order(HVG.df$bio, decreasing = TRUE),]
genes <- rownames(HVG.df)[1:1000]
l.genes <- list()
for(i in 1:length(datasets)){l.genes[[i]] <- genes}


# List of normalized count matrices
datasets <- lapply(sce.list, function(n){
  t(as.matrix(logcounts(n)[genes,]))
})

# Data integration
integrated.data <- scanorama$integrate(datasets, l.genes)
```

This doesn't seem to work.
So I'll try the Seurat implementation instead.

```{r}
library(Seurat)

# List of highly variable genes
# List of genes
HVG.genes <- lapply(sce.list, function(n){
    HVG <- trendVar(n, use.spikes = FALSE)
    decomposeVar(n, HVG)
})
  
HVG.df <- do.call("combineVar", HVG.genes)
HVG.df <- HVG.df[order(HVG.df$bio, decreasing = TRUE),]
genes <- rownames(HVG.df)[1:1000]

# SCE to Seurat objects
sce.list.seurat <- lapply(sce.list, function(n){
  as.Seurat(n, counts = "counts", data = "logcounts")
})

# Normalize and find variable features for each sample
#for (i in 1:length(sce.list.seurat)) {
#    sce.list.seurat[[i]] <- NormalizeData(sce.list.seurat[[i]], verbose = FALSE)
#    sce.list.seurat[[i]] <- FindVariableFeatures(sce.list.seurat[[i]], selection.method = "vst", 
#        nfeatures = 1000, verbose = FALSE)
#}

# Find anchors
anchors <- FindIntegrationAnchors(object.list = sce.list.seurat, dims = 1:30, 
                                  k.filter = 30, anchor.features = genes)

# Integration
integrated <- IntegrateData(anchorset = anchors, dims = 1:30)

# Visualization
DefaultAssay(integrated) <- "integrated"
integrated <- ScaleData(integrated, verbose = FALSE)
integrated <- RunPCA(integrated, npcs = 30, verbose = FALSE)
integrated <- RunTSNE(integrated, verbose = FALSE)

p.tsne.seurat.1 <- ggplot(data.frame(tSNE1 = integrated@reductions$tsne@cell.embeddings[,1],
                  tSNE2 = integrated@reductions$tsne@cell.embeddings[,2],
                  clusters = as.factor(integrated@meta.data$Tissue))) + 
  geom_point(aes(tSNE1, tSNE2, colour = clusters)) +
  scale_colour_manual(values = metadata(all_sce)$colour_vector)
ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/tsne_Seurat.wSMG.k30.pdf", 
       p.tsne.seurat.1, width = 12, height = 7)

# We now remove the SMG cells and run the pipeline with default values
sce.list.seurat <- sce.list.seurat[!grepl("SMG", names(sce.list.seurat))]

anchors <- FindIntegrationAnchors(object.list = sce.list.seurat, dims = 1:30, anchor.features = genes)

# Integration
integrated <- IntegrateData(anchorset = anchors, dims = 1:30)

# Visualization
DefaultAssay(integrated) <- "integrated"
integrated <- ScaleData(integrated, verbose = FALSE)
integrated <- RunPCA(integrated, npcs = 30, verbose = FALSE)
integrated <- RunTSNE(integrated, verbose = FALSE)

p.tsne.seurat.2 <- ggplot(data.frame(tSNE1 = integrated@reductions$tsne@cell.embeddings[,1],
                  tSNE2 = integrated@reductions$tsne@cell.embeddings[,2],
                  clusters = as.factor(integrated@meta.data$Tissue))) + 
  geom_point(aes(tSNE1, tSNE2, colour = clusters)) +
  scale_colour_manual(values = metadata(all_sce)$colour_vector)
ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/tsne_Seurat.woSMG.defaults.pdf", 
       p.tsne.seurat.2, width = 12, height = 7)

# Now with the Seurat normalization and feature selection
for (i in 1:length(sce.list.seurat)) {
    sce.list.seurat[[i]] <- NormalizeData(sce.list.seurat[[i]], verbose = FALSE)
    sce.list.seurat[[i]] <- FindVariableFeatures(sce.list.seurat[[i]], selection.method = "vst", 
        nfeatures = 1000, verbose = FALSE)
}

anchors <- FindIntegrationAnchors(object.list = sce.list.seurat, dims = 1:30)

# Integration
integrated <- IntegrateData(anchorset = anchors, dims = 1:30)

# Visualization
DefaultAssay(integrated) <- "integrated"
integrated <- ScaleData(integrated, verbose = FALSE)
integrated <- RunPCA(integrated, npcs = 30, verbose = FALSE)
integrated <- RunTSNE(integrated, verbose = FALSE)

p.tsne.seurat.3 <- ggplot(data.frame(tSNE1 = integrated@reductions$tsne@cell.embeddings[,1],
                  tSNE2 = integrated@reductions$tsne@cell.embeddings[,2],
                  clusters = as.factor(integrated@meta.data$Tissue))) + 
  geom_point(aes(tSNE1, tSNE2, colour = clusters)) +
  scale_colour_manual(values = metadata(all_sce)$colour_vector)
ggsave("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/BatchCorrectionValidation/tsne_Seurat.woSMG.defaults.SeuratNormHVG.pdf", p.tsne.seurat.3, width = 12, height = 7)
```

Next we will try scmap.

Next, we will try harmony.