---
title: "Figure 2: Similarity analysis between BE and other tissues"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script focuses on the similarity analysis between BE and other tissues: SCJ, GOJ, GC, SMG and D2.

# Read in the data

```{r libraries, message=FALSE}
library(scran)
library(scater)
library(DropletUtils)
library(openxlsx)
library(Rtsne)
library(pheatmap)
library(cowplot)
library(RColorBrewer)
library(edgeR)
library(ape)
library(viridis)
library(umap)
source("../Analysis/Functions/auxiliary.R")
sce <- readRDS("../../../Dropbox (Personal)/Oesophagus_single_cell/All_corrected_sce.rds")

# Exclude contaminating cells
sce <- sce[,sce$include]
```

# Plot tsne

Here, we visualize the batch-corrected counts using tsne.

```{r}
# Batch correction
sce.list <- split.sce(sce, unique(sce$Sample), colData.name = "Sample")
corrected <- batch.correction(sce.list)
sce <- do.call("cbind", sce.list)

# Compute new tSNE
set.seed(11111)
tsne <- Rtsne(t(corrected), pca = FALSE, perplexity = 250)
reducedDims(sce)$TSNE <- tsne$Y

# Visualize tsne
p.all.cells <- ggplot(data.frame(tsne1 = reducedDims(sce)$TSNE[,1],
                  tsne2 = reducedDims(sce)$TSNE[,2],
                  tissue = colData(sce)$Tissue)) + 
  geom_point(aes(tsne1, tsne2), colour = "black", size = 1) + 
  geom_point(aes(tsne1, tsne2, colour = tissue), size = 0.5) + 
  scale_color_manual(values = metadata(sce)$colour_vector) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "grey"))

ggsave("../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Figure_2/All_cells.pdf", p.all.cells, width = 7, height = 5)
```

# Plot circular dendrogram of all clusters

To estimate the similarity between all cell types, we compute an averaged expression profile for each cell-type and perform hierarchical cluster analysis.

```{r}
# We will perform this analysis on the corrected and uncorrected counts (uncorrected in supplements)
clusters <- paste(colData(sce)$Tissue, colData(sce)$Tissue_cluster, sep = "_")
mat <- matrix(data = NA, ncol = length(unique(clusters)), nrow = nrow(corrected))
colnames(mat) <- unique(clusters)

for(i in unique(clusters)){
  mat[,i] <- rowMeans(corrected[,clusters == i])
}

# Rename the matrix to contain the actual cell-type labels
for(i in 1:ncol(mat)){
  colnames(mat)[i] <- unique(sce$cell_type[sce$Tissue == sub("_[0-9]*$", "", colnames(mat)[i]) &
                                 sce$Tissue_cluster == sub("^[A-Z2]*_", "", colnames(mat)[i])])
}

# Calculate euclidean distance on corrected counts
dend <- hclust(dist(t(mat), method = "euclidean"), method = "ward.D2")
dev.off()
pdf("../../../Dropbox (Personal)/Oesophagus_single_cell/Results/Figures/Figure_2/Tree_corrected_counts.pdf", width = 10, height = 10)
plot(as.phylo(dend), type = "fan", 
     tip.color = metadata(sce)$colour_vector[sub("_[0-9]*$", "", unique(clusters))])
dev.off()
```