---
title: "Characterisation of pure normal tissues"
author: "Juliane Perner"
date: "05/04/2018"
output: html_document
---

###########################
# Preliminaries
###########################

```{r setup, include=FALSE}
	# Rscript -e "library(rmarkdown); render('ByTissueClustering.Rmd')"
	knitr::opts_chunk$set(	echo=TRUE,
				results='asis',
				fig.align='center')
```

##############
## Load libraries and functions
##############

```{r libraries, message=FALSE}
library(ggplot2)
library(pheatmap)
library(Rtsne)
library(scran)
library(cowplot)
library(RColorBrewer)
library(Matrix)
library(dynamicTreeCut)
library(reshape2)
library(cluster)
source("../Functions/auxiliary.R")
```

##############
## Set variables and paths to data
##############

I'm only looking into NE, SMG, GC and D2 taking all data we have for these tissues, regardless of whether the patient had Barrett's or not. Samples from Patient 4 and the gland and duct samples from Patient 6 were excluded because only cultured cells were sequenced. Patient 5 was excluded because of ambiguous pathology and high normal esophagus content in both GOJ and GC samples. 

```{r pathsAndglobalVars}
# input folders, tissues, patients
patientFolders <- "/Users/nils/Dropbox (Cambridge University)/Oesophagus_single_cell/"
tissues <- c("NE", "SMG", "GOJ", "GC", "D2", "BE", "SCJ")
patients <- c("Patient1", "Patient2", "Patient3", "Patient6", "Patient7", "Patient8", "Patient9", "Patient10", "Patient11")

# reference patient ID for batch correction for each tissue
refPatientID <- c( NE="Patient2", 
		SMG="Patient11",
		GOJ="Patient2",
		GC="Patient2", 
		D2="Patient9",
		BE="Patient7",
		SCJ="Patient3")

# set patient categories
patientCats <- c(`1`="Biopsy", 
		`2`="Biopsy", 
		`3`="Barrett's",
		`6`="Donor", 
		`7`="Barrett's", 
		`8`="Biopsy", 
		`9`="Barrett's",
		`10`="Donor",
		`11`="Donor")

# set graphic parameters
patientCols <- brewer.pal(length(patients), "Set1")
names(patientCols) <- gsub("Patient", "", patients)
```

############################
# Load and batch-correct data
############################

Here, I load the available samples for each tissue type and batch-correct each tissue type across patient. First, only genes present in all cells and expressed in at least one cell of a tissue are retained. The top-1000 most highly variable genes are obtained by applying decomposeVar to the log2-transformed data. Batch-correction across patients is performed using mnnCorrect and the highly variable genes.

```{r buildTissueRef}
# find all files per tissue and selected patients
files_byTissue <- sapply(tissues, function(x) list.files(patientFolders, pattern=paste0("*_", x, "_norm.rds"), recursive=TRUE, full.name=TRUE))
files_byTissue <- sapply(files_byTissue, function(x) grep(paste0(patients, collapse="|"), x, value=TRUE))

# reorder for use in reference in batch correction
files_byTissue <- sapply(names(files_byTissue), function(tissue) {
	files <- files_byTissue[[tissue]]
	rp <- refPatientID[[tissue]]
	if(any(grepl(rp, files))) {
		c(grep(rp, files, value=TRUE), grep(rp, files, value=TRUE, invert=TRUE)) 
	}else{ files }})

# go through each tissue
tsne_byTissue <-list() # tsne result based on batch corrected data and highly variable genes
hgv_byTissue <- list()	# highly variable genes on all samples from same tissue
bcdata_byTissue <- list() # batch corrected data
normdata_byTissue <- list() # QC-passed data
for(tissue in tissues){
	# load data
	normdata_byTissue[[tissue]] <- loadData(files_byTissue[[tissue]])

	# filter CD45+ or VIM+ cells
	#normdata_byTissue[[tissue]] <- filterData(normdata_byTissue[[tissue]], cd45=TRUE, vim=TRUE)

	# find highly variable genes
	hgv_byTissue[[tissue]] <- getHVGs(normdata_byTissue[[tissue]])

	# batch correct if more than one sample
	if(length(files_byTissue[[tissue]])>1){
		# batch correct
		bcdata_byTissue[[tissue]] <- batchCorrect(normdata_byTissue[[tissue]], hgv_byTissue[[tissue]])
		# reset colnames and combine
		for(i in 1:length(bcdata_byTissue[[tissue]]$corrected)) {
			colnames(bcdata_byTissue[[tissue]]$corrected[[i]]) <- colnames(normdata_byTissue[[tissue]][[i]])
		}
		bcdata_byTissue[[tissue]] <- do.call("cbind", bcdata_byTissue[[tissue]]$corrected)
	}else{
		# just log transform
		bcdata_byTissue[[tissue]] <- log2( as.matrix(normdata_byTissue[[tissue]][[1]])[hgv_byTissue[[tissue]], ] +1 )
	}

	# perform tsne
	set.seed(123)
	tsne_byTissue[[tissue]] <- Rtsne( t(bcdata_byTissue[[tissue]]) )
}
```

############################
# QC by tissue
############################

We expect similar cellular composition for samples of the same tissue. Hence, in the t-SNE plot all cell groups should be supported by several patients. Further, cells with low quality should not stand out as a separate group. I am using the raw data to calculate different QC statistics.

##############
## Prepare QC data
##############

```{r qcSummary}
# find all relevant raw data files
files <- list.files(patientFolders, pattern="*matrix.mtx", recursive=TRUE, full.name=TRUE)
files <- grep(paste0(patients, collapse="|"), files, value=TRUE)
files <- grep(paste0(tissues, collapse="|"), files, value=TRUE)

# process one file at a time ---- FROM NILS QC_NORM; BETTER TO GET STATS FROM THERE?
qcHelper <- function(file){
	# load data
	input <- readMM(file)
	prefix <- paste0(grep("Patient|SIGA|HCAT", strsplit(file, "/")[[1]], value=TRUE), collapse="_")
	colnames(input) <- paste(prefix, seq(1,ncol(input)), sep="_")
	genes <- read.table(gsub("matrix.mtx", "genes.tsv", file))
	rownames(input) <- genes[,2]

	# get library size per cell
	libsize <- colSums(input)

	# get number of expressed genes by cell
	numexpr <- apply(input>0, 2, sum)

	# get percent of ribosomal rna
	ribogenes <- readLines("ribogenes.txt")
	ribogenes <- row.names(input)[apply(sapply(ribogenes, grepl, x=row.names(input)), 1, any)]
	ribopercent <- colSums(input[ribogenes,])/libsize
		
	# get percent of mitochondria rna
	mitogenes <- grepl("^MT-", row.names(input))
	mitopercent <- colSums(input[mitogenes,])/libsize
	
	return(cbind(libsize, numexpr, ribopercent, mitopercent))
}
qcstats <- sapply(files, qcHelper)
qcstats <- do.call("rbind", qcstats)
```

##############
## Plot QC 
##############

```{r plotQC, fig.width=25, fig.height=25}
# for shared legends
qcstats <- qcstats[unlist(sapply(bcdata_byTissue, colnames)),]
libsize_lim <- summary(log10(qcstats[,"libsize"]))[c("Min.", "Max.")]
numexpr_lim <- summary(qcstats[,"numexpr"])[c("Min.", "Max.")]
ribo_lim <- summary(qcstats[,"ribopercent"])[c("Min.", "Max.")] *100
mito_lim <- summary(qcstats[,"mitopercent"])[c("Min.", "Max.")] *100

# make all plots for all tissues
plots <- vector("list", length(tissues))
names(plots) <- tissues
for( tissue in tissues ) {
	# make data frame with QC data for plotting
	#samples <- sub("_[0-9].*", "", colnames(bcdata_byTissue[[tissue]]))
	#samples <- factor(sampleToPatient[samples])
	samples <- factor(gsub("Patient", "", sub("_.*", "", colnames(bcdata_byTissue[[tissue]]))))
	cells <- colnames(bcdata_byTissue[[tissue]])
	toplot <- data.frame( tsne1=tsne_byTissue[[tissue]]$Y[,1], 
				tsne2=tsne_byTissue[[tissue]]$Y[,2],
				Patient=factor(samples, levels=names(patientCols)),
				PatientClass=factor(patientCats[samples], levels=unique(patientCats)))
	toplot <- cbind(toplot, qcstats[cells,])

	# plot the tsne plot with different coloring
	basePlot <- ggplot(toplot, aes(x=tsne1, y=tsne2)) +
			background_grid(major="none", minor="none") +
			ggtitle(tissue) + xlab("t-SNE1") + ylab("t-SNE2") + 
			theme(	legend.title=element_text(size=10, face="bold"), 
				legend.text=element_text(size=10),
				legend.position="none") 

	# highlighting the patient
	plots[[tissue]]$patient <- basePlot + geom_point(aes(colour=Patient), size=.25) + 
			scale_color_manual(values=patientCols) +
			guides(colour=guide_legend(override.aes=list(size=2)))

	plots[[tissue]]$patientClass <- basePlot + geom_point(aes(colour=PatientClass), size=.25) + 
			scale_color_brewer(palette="Set2") +
			guides(colour=guide_legend(override.aes=list(size=2)))
	
	# highlighting the libsize
	plots[[tissue]]$libsize <- basePlot + 
			geom_point(aes(colour=log10(libsize)), size=.25) + 
			scale_color_gradientn(colours=brewer.pal(5, "YlGnBu"), 
						limits=libsize_lim, 
						name="log10(library size)")
	
	# highlighting the numexpr
	plots[[tissue]]$numexpr <- basePlot + 
			geom_point(aes(colour=numexpr), size=.25) + 
			scale_color_gradientn(colours=brewer.pal(5, "YlGnBu"), 
						limits=numexpr_lim, 
						name="Expressed genes")
	
	# highlighting the ribopercent
	plots[[tissue]]$ribopercent <- basePlot + 
			geom_point(aes(colour=ribopercent*100), size=.25) + 
			scale_color_gradientn(colours=brewer.pal(5, "Spectral"), 
						limits=ribo_lim, 
						name="Ribosomal")
	
	# highlighting the mitopercent
	plots[[tissue]]$mitopercent <- basePlot + 
			geom_point(aes(colour=mitopercent*100), size=.25) + 
			scale_color_gradientn(colours=brewer.pal(5, "Spectral"), 
						limits=mito_lim, 
						name="Mitochondrial")
}

# add legends
legends <- list()
for(i in 1:length(plots[["GC"]])) {
	legends[[i]] <- get_legend(plots[["GC"]][[i]] + theme(legend.position="right"))
}
names(legends) <- names(plots[["GC"]])

# plot everything into one Figure plot; TODO
what <- names(plots[["NE"]])
n <- length(plots)
lines <- sapply(what, function(x) paste0( sapply(tissues, function(y) paste0("plots[[\"", y, "\"]][[\"", x, "\"]]")), collapse=", "))
lines <- sapply(what, function(x) paste(lines[x], paste0("legends[[\"", x, "\"]]"), sep=", "))
funcText <- paste0("plot_grid( ", paste0(lines, collapse=", "),
		", labels=c(\"A\", rep(\"\", n), \"B\", rep(\"\", n), \"C\", rep(\"\", n), \"D\", rep(\"\", n), \"E\", rep(\"\", n)), ncol=n+1, rel_widths=c(rep(1,n),0.5))")
eval( parse(text=funcText) )
```

############################
# Clustering by tissue
############################

For clustering cells into distinct populations, I'm performing hierarchical clustering using the Ward.D2 agglomeration method and the distance measure based on Spearman's correlation coefficient on the batch-corrected data. The dendrogram is then cut using dynamicTreeCut with various combinations of minimum cluster size and deepSplit to vary the granularity of the clustering. The final clustering solution has maximal average silhouette score. For quality control, each cluster should be made up of cells from different patients and each patient should contribute to each cluster.

##############
## Clustering
##############

```{r perTissueCluster}
# possible parameters for cutreeDynamic
mcl <- c(20, 50, 100, 150, 200)
ds <- 0:4

# perform hierarchical clustering with dynamic tree cut on each tissue
clusters_byTissue <- vector("list", length(tissues))
names(clusters_byTissue) <- tissues
for( tissue in tissues ){
	# distance matrix based on spearman correlation
	d <- as.dist(sqrt((1 - cor(bcdata_byTissue[[tissue]], method="spearman"))/2))
	# hierarchical clustering
	hc <- hclust(d, method="ward.D2")
	# dynamic tree cut
	bestscore <- 0
	for(cur_mcl in mcl){
		for(cur_ds in ds){
			tmp <- cutreeDynamic(dendro=hc, 
						distM=as.matrix(d), 
						minClusterSize=cur_mcl, 
						deepSplit=cur_ds, 
						verbose=0)
			sil <- summary(silhouette(tmp, dist=d))$avg.width
			if( sil>=bestscore & !any(tmp==0) ){
				bestscore <- sil
				clusters_byTissue[[tissue]] <- tmp
				names(clusters_byTissue[[tissue]]) <- hc$labels
			}
		}
	}
}
```

##############
## QC of cluster solutions
##############

```{r plotClusterQC, fig.width=20, fig.height=7, message=FALSE}
# helper function to tabulate and get fractions (either cluster by patient or patient by cluster)
fracHelper <- function(x, y, clustername, samplename){
	# table x by y	
	tab <- tapply(x, y, table)

	# make fractions
	tab <- lapply(tab, function(cur_tab) cur_tab/sum(cur_tab))

	# convert to data frame for plotting
	tab <- melt(tab)
	# if you one patient add column
	#if(length(unique(x))==1){
	#	tab$Patient <- sampleToPatient[as.character(unique(x))]
	#}else{
	#	tab$Patient <- sampleToPatient[as.character(tab[,samplename])]
	#}
	tab$Cluster <- factor(tab[, clustername], levels=0:max(tab[, clustername]))
	return(tab)
}

# for each patient, check fraction of cells in each cluster
fracplots1 <- vector("list", length(tissues))
names(fracplots1) <- tissues
for( tissue in tissues ) {

	# table cells by cluster and sample
	#samples <- sub("_[0-9].*", "", colnames(bcdata_byTissue[[tissue]]))
	samples <- factor(gsub("Patient", "", sub("_.*", "", colnames(bcdata_byTissue[[tissue]]))))
	tab_cl_sample <- fracHelper(x=clusters_byTissue[[tissue]], y=samples, clustername="Var1", samplename="L1")

	# barplot of fractions
	fracplots1[[tissue]] <- ggplot(tab_cl_sample, aes(x=L1, y=value, fill=Cluster)) +
				geom_bar(stat="identity") +
				background_grid(major="none", minor="none") +
				ggtitle(tissue) + xlab("Patient") + ylab("Fraction of cells") + 
				theme(	legend.title=element_text(size=10, face="bold"),
					legend.text=element_text(size=10)) +
				scale_fill_manual(values=c(brewer.pal(8, "Dark2"), rev(brewer.pal(8, "Set1"))))
}

# for each patient, check fraction of cells in each cluster
fracplots2 <- vector("list", length(tissues))
names(fracplots2) <- tissues
for( tissue in tissues ) {

	# table cells by cluster and sample
	#samples <- sub("_[0-9].*", "", colnames(bcdata_byTissue[[tissue]]))
	samples <- factor(gsub("Patient", "", sub("_.*", "", colnames(bcdata_byTissue[[tissue]]))))
	tab_cl_sample <- fracHelper(y=clusters_byTissue[[tissue]], x=samples, clustername="L1", samplename="Var1")

	# barplot of fractions
	fracplots2[[tissue]] <- ggplot(tab_cl_sample, aes(x=Cluster, y=value, fill=factor(Var1))) +
				geom_bar(stat="identity") +
				background_grid(major="none", minor="none") +
				ggtitle(tissue) + xlab("Cluster") + ylab("Fraction of cells") + 
				theme(	legend.title=element_text(size=10, face="bold"),
					legend.text=element_text(size=10)) +
				scale_fill_manual(values=patientCols) + 
				guides(fill=guide_legend(title="Patient"))
}
n <- length(fracplots1)
funcText <- paste0("plot_grid( ", paste(sapply(names(fracplots1), function(x) paste0("fracplots1[[\"", x ,"\"]]")), collapse=",") , ",",
			paste(sapply(names(fracplots2), function(x) paste0("fracplots2[[\"", x ,"\"]]")), collapse=","), 
		", labels=c(\"A\", rep(\"\", n), \"B\", rep(\"\", n)), nrow=2)")
eval( parse(text=funcText) )
```

##############
## Plot cluster solutions
##############

```{r plotPerTissueCluster, fig.width=20, fig.height=8, message=FALSE}
# make tSNE plots
tsneplots <- vector("list", length(tissues))
names(tsneplots) <- tissues
for( tissue in tissues ) {
	# make data frame with QC data for plotting
	toplot <- data.frame( tsne1=tsne_byTissue[[tissue]]$Y[,1], 
				tsne2=tsne_byTissue[[tissue]]$Y[,2],
				Cluster=factor(clusters_byTissue[[tissue]]))

	# cluster colors and labels
	cols <- c("black", brewer.pal(8, "Dark2"), rev(brewer.pal(8, "Set1")))
	names(cols) <- 0:(length(cols)-1)
	cl_tab <- table(clusters_byTissue[[tissue]])
	cols_lab <- min(as.numeric(names(cl_tab))):max(as.numeric(names(cl_tab)))
	cols_lab <- sapply(cols_lab, function(x) paste0(x, " (n=", cl_tab[as.character(x)], ")"))

	# plot the tsne plot with different coloring
	tsneplots[[tissue]] <- ggplot(toplot, aes(x=tsne1, y=tsne2)) +
			geom_point(aes(colour=Cluster), size=.5) +
			background_grid(major="none", minor="none") +
			ggtitle( paste0(tissue, " (n=", nrow(toplot), ")") ) + 
			xlab("t-SNE1") + ylab("t-SNE2") + 
			theme(	legend.title=element_text(size=10, face="bold"), 
				legend.text=element_text(size=10)) +
			scale_color_manual(values=cols, labels=cols_lab) + 
			guides(colour=guide_legend(override.aes=list(size=2)))
}

# plot everything into one Figure plot
n <- length(tsneplots)
funcText <- paste0("plot_grid( ", paste(sapply(names(tsneplots), function(x) paste0("tsneplots[[\"", x ,"\"]]")), collapse=",") , ", nrow=2)")
eval( parse(text=funcText) )
```

##############
## Plot known markers
##############

```{r prepPlotHandpickedMarkers}
# list of genes whose expression we would like to check in each tissue
genesToPlot <- list( GC=c("MUC5AC", "TFF1", "GKN1", "GKN2", "MUC6", "TFF2", "PGA3", "PGA4", "PGA5", "PGC", "GIF", "GAST", "CHGA", "LIPF", "LYZ", "VIM"),
			NE=c("KRT14", "TP63", "KRT5", "KRT13", "KRT4", "KRT3", "KRT7", "KRT6C", "KRT6B", "KRT8", "KRT19", "VIM"),
			D2=c("SOX9", "ALPI", "GHRL", "GRHL1", "GRHL2", "GRHL3", "CCK", "SCT", "TRPM5", "DLL1", "OLFM4", "HES1", "BMI1", "DCLK1", "GFI1", "MSI1", "PROM1", "LYZ", "LGR5", "ASCL2", "TFF3", "REG4", "CHGA", "CA1", "CA2", "KRT20", "CEACAM1", "VIM"))

# geneID to HGNC map
geneMap <- read.table(paste0(patientFolders, "Patient3/Data/SIGAA5_NE/genes.tsv"), stringsAsFactors = FALSE)
rownames(geneMap) <- geneMap[,1]

# plot
plotlist <- list()
for(tissue in tissues){
	# transform data for plotting
	data <- sapply(normdata_byTissue[[tissue]], function(x) log2(x+1))
	data <- do.call("cbind", data)

	# go through all genes
	tmp <- list()
	for(g in genesToPlot[[tissue]]){
		# get the ensembl id for that gene
		e <- geneMap[geneMap[,2]%in%g, 1]
	
		# reduce gene list to genes present in the data
		if( e %in% row.names(data) ){	
			# make df for plotting
			toplot <- data.frame( tsne1=tsne_byTissue[[tissue]]$Y[,1], 
						tsne2=tsne_byTissue[[tissue]]$Y[,2],	
						norm_expr=data[e,],
						gene=g)
			tmp[[g]] <- ggplot(toplot, aes(x=tsne1, y=tsne2)) +
						geom_point(aes(color=norm_expr), size=.5) +
						background_grid(major="none", minor="none") +
						xlab("t-SNE1") + ylab("t-SNE2") + ggtitle(g) +
						theme( axis.line=element_blank(),
							axis.text.x=element_blank(),
							axis.text.y=element_blank(),
							axis.ticks=element_blank(),
							axis.title.x=element_blank(),
							axis.title.y=element_blank(),
							panel.background=element_blank(),
							panel.border=element_blank(),
							panel.grid.major=element_blank(),
							panel.grid.minor=element_blank(),
							plot.background=element_blank()) +
						scale_color_gradientn(colours=rev(colorRampPalette(brewer.pal(9, "YlGnBu"))(100))) 
		}
	}
	if(length(tmp)>0){
		#legend <- get_legend(tmp[[1]] + theme(legend.position="right"))
		funcText <- paste0("plot_grid(", 
					paste0("tmp[[", 1:length(tmp), "]]", collapse=", 
					"), ", ncol=4)")
		plotlist[[tissue]] <- eval( parse(text=funcText) )
	}
}		
```

##############
### GC
##############
```{r plotHandpickedMarkers1, fig.width=20, fig.height=15, message=FALSE}
print(plotlist[["GC"]])
```

##############
### NE
##############
```{r plotHandpickedMarkers2, fig.width=15, fig.height=10, message=FALSE}
print(plotlist[["NE"]])
```

##############
### D2
##############
```{r plotHandpickedMarkers3, fig.width=15, fig.height=20, message=FALSE}
print(plotlist[["D2"]])
```

#############
# Marker gene detection 
#############

In each tissue, I'm looking for genes that are specifically enriched in a cluster compared to all other clusters. I use the normalised, log2-transformed but not batch-corrected data for each tissue and the findMarkers function with the patient as blocking term. Genes with a FDR < 0.1 (based on combined p-values across contrasts) and a log-foldchange of > 1 over all contrasts are selected. 

```{r markerGenes}
# marker plots
markers_byTissue <- vector("list", length(tissues))
names(markers_byTissue) <- tissues
for( tissue in tissues ) {
        # merge data
        genes <- lapply(normdata_byTissue[[tissue]], row.names)
        genes <- table(unlist(genes))
        genes <- names(genes)[genes>=length(normdata_byTissue[[tissue]])]
        prepData <- sapply(normdata_byTissue[[tissue]], function(x){x[genes,]})
        prepData <- do.call("cbind", prepData)

        # subselect expressed genes and log-transform
        prepData <- prepData[rowMeans(prepData)>0.1, ]
        prepData <- log2(prepData+1)

        # get marker genes
        samples <- intersect(colnames(prepData), names(clusters_byTissue[[tissue]]))
	patient <- sapply(strsplit(samples, "_"), "[", 1)
	design <- model.matrix(~patient)
        markerres <- findMarkers(prepData[,samples], clusters=clusters_byTissue[[tissue]][samples], design=design)
        markers_byTissue[[tissue]] <- lapply(markerres, function(x) {
				x[x$FDR<0.1 & apply(x[grepl("logFC", names(x))]>0, 1, all), ] })
}
```

```{r plotMarkerGenesFunction}
# plot the heatmap of markers
plotMarkersTissue <- function(tissue){
	# set possible cluster colors 
	cols <- c("black", brewer.pal(8, "Dark2"), rev(brewer.pal(8, "Set1")))
	names(cols) <- 0:(length(cols)-1)

	# select only the genes among the top 100 per cluster
	topmarkers <- sapply(markers_byTissue[[tissue]], function(x) x[x$Top<100, "Gene"])
	if(any(duplicated(unlist(topmarkers)))) warning("duplicated marker genes")
	tm <- unique(unlist(topmarkers))	

	# transform data for plotting
	data <- sapply(normdata_byTissue[[tissue]], function(x) log2(x+1))
	data <- do.call("cbind", data)

	# mean-scale data by gene
	toplot <- data - rowMeans(data)

	# reorder genes and samples by cluster
	toplot <- toplot[tm, names(sort(clusters_byTissue[[tissue]]))]

	# map gene names
	row.names(toplot) <- geneMap[tm, 2]
	
	# plot 
	minmax <- max(abs(toplot))
	df_col <- data.frame(cluster=factor(clusters_byTissue[[tissue]]), row.names=names(clusters_byTissue[[tissue]]))
	df_row <- data.frame(cluster=factor(rep(names(topmarkers), sapply(topmarkers, length))), row.names=geneMap[tm, 2])
	pheatmap(toplot, cluster_cols=FALSE, cluster_rows=FALSE, 
		annotation_col=df_col,
		annotation_row=df_row,
		annotation_colors=list(cluster=cols[levels(df_col$cluster)]),
		gaps_row=cumsum(table(df_row$cluster)),
		gaps_col=cumsum(table(df_col$cluster)),
		show_colnames=FALSE, show_rownames=FALSE,
		breaks=seq(-minmax, minmax, length.out=51),
        	color=colorRampPalette(c("navy", "grey94", "firebrick3"))(50), main=tissue)

	return(df_row)
}
```

```{r plotMarkerGenes1, fig.width=7, fig.height=10}
m <- plotMarkersTissue("GC")
write.csv(m, file="markers_GC.csv")
```

```{r plotMarkerGenes2, fig.width=14, fig.height=10}
m <- plotMarkersTissue("NE")
write.csv(m, file="markers_NE.csv")
```

```{r plotMarkerGenes3, fig.width=5, fig.height=5}
m <- plotMarkersTissue("D2")
write.csv(m, file="markers_D2.csv")
```

```{r plotMarkerGenes4, fig.width=5, fig.height=5}
m <- plotMarkersTissue("SMG")
write.csv(m, file="markers_SMG.csv")
```

##############
## Get differentially expressed genes between normal tissues
##############

```{r goEnr, message=FALSE, warning=FALSE}
# functions for GO enrichment
source("goEnr.R")
# mean expression across all cells
rm <- rowMeans(prepData)

markers_all <- unlist(markers_byTissue, recursive=FALSE)

res <- list()
for(i in 1:length(markers_all)){
	# get markers
	geneselect <- intersect(names(rm), markers_all[[i]]$Gene)

	# make background with similar expression
	universe <- prepUniverse(rm, geneselect)
	
	# prep input
	newGeneList <- makeGeneList(universe, geneselect)
	
	# run GO-enrichment
	testall <- runEnrichment(newGeneList, topnodes=50)
	res[[i]] <- testall
}

goenr <- lapply(res, function(x) x[ x$Fisher.classic<0.01 & x$Fisher.elim<0.05 , ])
names(goenr) <- names(markers_all)
```

```{r saveall, echo=FALSE}
save(list=c("clusters_byTissue", "normdata_byTissue", "markers_byTissue", "tsne_byTissue", "goenr"), file="byTissue.RData")
```
