---
title: "Trajectory analysis"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, 
        encoding = encoding, output_file = '../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Trajectories/Trajectory_BE.html') })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(reticulate)
use_python("/usr/bin/python")
mpl = reticulate::import("matplotlib")
mpl$use('TkAgg')
```

This script performs trajectory analysis on the BE and GC samples

## Read in batch corrected data

```{r, message=FALSE, warning=FALSE}
library(scran)
library(Rtsne)
library(ggplot2)
library(viridis)
library(monocle)
library(plot3Drgl)
library(umap)
library(vegan)
library(pheatmap)
library(openxlsx)
suppressMessages(source("../Functions/auxiliary.R"))
sce <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/All_corrected_sce.rds")

# Select only BE
sce <- sce[,colData(sce)$Tissue == "BE"]
sce <- sce[Matrix::rowMeans(counts(sce)) > 0.1,]

# Split sce object
sce_single <- split.sce(sce, groups = unique(colData(sce)$Sample), colData.name = "Sample")

corrected <- batch.correction(sce_single)

sce <- do.call("cbind", sce_single)

colnames(corrected) <- paste(colData(sce)$Barcode, colData(sce)$Patient, colData(sce)$Tissue, sep = "_")
```

# Try different visualization techniques

## PCA

```{r}
# Batch corrected data already come as pca
ggplot(data.frame(PC1 = corrected[1,],
                  PC2 = corrected[2,],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = PC1, y = PC2, colour = patient))

# Patient 7 contains NE cells - remove them for BE analysis
sce <- sce[,corrected[1,] > -0.01]
corrected <- corrected[,corrected[1,] > -0.01]

ggplot(data.frame(PC1 = corrected[1,],
                  PC2 = corrected[2,],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = PC1, y = PC2, colour = patient))

ggplot(data.frame(PC1 = corrected[1,],
                  PC2 = corrected[2,],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = PC1, y = PC2, colour = cluster))

# Visualize OLFM4
ggplot(data.frame(PC1 = corrected[1,],
                  PC2 = corrected[2,],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = PC1, y = PC2, colour = OLFM4)) + scale_colour_viridis()
```

## tSNE

```{r}
set.seed(12345)
tsne <- Rtsne(t(corrected), pca = FALSE)

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = cluster))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = patient))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = tsne1, y = tsne2, colour = OLFM4)) + scale_colour_viridis()

```

# Diffusionmap

```{r}
library(destiny)

dm <- DiffusionMap(t(corrected), k = 20)

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = DC1, y = DC2, color = patient))

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = DC1, y = DC2, color = cluster))

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = DC1, y = DC2, colour = OLFM4)) + scale_colour_viridis()

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  RHOF = logcounts(sce)[rowData(sce)$Symbol == "RHOF",])) +
  geom_point(aes(x = DC1, y = DC2, colour = RHOF)) + scale_colour_viridis()
```

# PHATE

```{r}
library(phateR)

tree.phate <- phate(t(corrected), gamma = 0)

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = phate1, y = phate2, colour = patient))

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = phate1, y = phate2, colour = cluster))

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = phate1, y = phate2, colour = OLFM4)) + scale_colour_viridis()

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "MUC2",])) +
  geom_point(aes(x = phate1, y = phate2, colour = OLFM4)) + scale_colour_viridis()
```

# UMAP

```{r}
umap.all <- umap(t(corrected))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = patient))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = cluster))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = OLFM4)) + scale_colour_viridis()
```

# Entropy

The stemID score is calculated based on the cell-specific entropy times the number of links between clusters.
We can therefore calculate the cell-specific entropy as a proxy for stemness.
The entropy has to be computed on the normalized counts.

```{r entropy}
# These functions are taken from the RaceID3_StemID2_class.R script
# 1. approach - Add pseudotcount of 0.1
cur_counts <- 2^(logcounts(sce)) - 1 + 0.1
probs   <- t(t(cur_counts)/apply(cur_counts,2,sum))
entropy <- -apply(probs*log(probs)/log(nrow(sce)),2,sum)

# Cluster with highest median entropy
clust.entropy <- vector(length = length(unique(colData(sce)$Tissue_cluster)))
names(clust.entropy) <- as.character(unique(colData(sce)$Tissue_cluster))
for(i in unique(colData(sce)$Tissue_cluster)){
  clust.entropy[as.character(i)] <- mean(entropy[as.numeric(colData(sce)$Tissue_cluster) == i])
}

ggplot(data.frame(score = as.numeric(clust.entropy),
                  clust.names = names(clust.entropy))) +
  geom_point(aes(clust.names, score)) + ylim(c(0.6, 0.9))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = cluster))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  entropy = entropy)) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  entropy = entropy)) + 
  geom_point(aes(x = phate1, y = phate2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  entropy = entropy)) + 
  geom_point(aes(x = DC1, y = DC2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  entropy = entropy)) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

# Correlation between stemness and OLFM4 expression
ggplot(data.frame(entropy = entropy,
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(entropy, OLFM4))

# Find marker genes for stem cells
cors <- t(apply(logcounts(sce), 1, function(n){
  cur_cor <- suppressWarnings(cor.test(n, entropy, method = "spearman"))
  c(cur_cor$estimate, cur_cor$p.value)
}))
cors <- as.data.frame(cors)
cors$BH <- p.adjust(cors$V2, method = "BH")
cors$Symbol <- rowData(sce)$Symbol[match(rownames(cors), rowData(sce)$ID)]

# Order correlations
cors.pos <- cors[cors$rho > 0 & cors$BH < 0.01,]
head(cors.pos[order(cors.pos$rho, decreasing = TRUE),], n = 100)

# PTMA is a pluripotency marker, Kolodziejczyk 2015, Lin 2011
# Lin 2011, Plos One also finds: TPT1, PTMA, NCL, NASP, and HSP90AB1
ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  PTMA = logcounts(sce)[rowData(sce)$Symbol == "PTMA",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = PTMA)) + scale_colour_viridis()

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  NCL = logcounts(sce)[rowData(sce)$Symbol == "NCL",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = NCL)) + scale_colour_viridis()
```

Cluster 9 and 3 appear to be stem cell like clusters. 
Here, I will perform differential expression between these and all other clusters.

# Differential expression analysis

```{r}
pheatmap(t(t(table(sce$Tissue_cluster, sce$Patient))/colSums(table(sce$Tissue_cluster, sce$Patient))),
         cluster_rows = FALSE, cluster_cols = FALSE)

cur_clusters <- sce$Tissue_cluster 
cur_clusters[cur_clusters == 9 | cur_clusters == 3] <- "stem"

# Differential expression analysis
for(i in c(2,5,6,7,8,10,11,12,13)){
  cur_sce <- sce[,cur_clusters == "stem" | cur_clusters == as.character(i)]
  cur_sce <- cur_sce[rowMeans(counts(cur_sce)) > 0,]
  cluster_test <- cur_clusters[cur_clusters == "stem" | cur_clusters == as.character(i)]

  mat <- matrix(data = NA, ncol = length(unique(paste(colData(cur_sce)$Patient, 
                                        cluster_test, sep = "_"))), 
                nrow = nrow(counts(cur_sce)))
  rownames(mat) <- rownames(counts(cur_sce))
  colnames(mat) <- unique(paste(colData(cur_sce)$Patient, 
                                cluster_test, sep = "_"))
  
  for(j in colnames(mat)){
    cur_patient <- unlist(strsplit(j, "_"))[1]
    cur_clust <- unlist(strsplit(j, "_"))[2]
    mat[,j] <- Matrix::rowSums(counts(cur_sce)[,colData(cur_sce)$Patient == cur_patient &
                                                  cluster_test == cur_clust]) 
  }
  
  # Perform differential testing
  y <- DGEList(counts=mat,
              group=sapply(colnames(mat), 
                            function(n){unlist(strsplit(n, "_"))[1]}))
  y <- calcNormFactors(y)
  design <- model.matrix(~0+sapply(colnames(mat), function(n){unlist(strsplit(n, "_"))[2]}))
  colnames(design) <- c(paste("Cluster", i, sep = ""), "stem")
  y <- estimateDisp(y,design)
  
  fit <- glmQLFit(y,design, robust = TRUE)
  qlf <- glmTreat(fit,coef=2, lfc = 0.5, 
                 contrast = eval(parse(text=paste("makeContrasts(Cluster", i, " - stem, levels = design)", sep = ""))))
  cur_markers <- topTags(qlf, n = nrow(qlf$table))$table

  # Save markers
  cur_out <- list()
  cur_out$Stem <- cur_markers[cur_markers$logFC < 0 & cur_markers$FDR < 0.1,]
  cur_out$Stem$Genename <- rowData(cur_sce)$Symbol[match(rownames(cur_out$Stem), rowData(cur_sce)$ID)]
  cur_out[[paste("Cluster", i, sep = "")]] <- cur_markers[cur_markers$logFC > 0 & cur_markers$FDR < 0.1,]
  cur_out[[paste("Cluster", i, sep = "")]]$Genename <- rowData(cur_sce)$Symbol[match(rownames(cur_out[[paste("Cluster", i, sep = "")]]), rowData(cur_sce)$ID)]

  # Save output
  write.xlsx(cur_out, paste("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Marker_genes/DE/Barrets_clusters/Cluster", i, 
                            "_vs_StemCells.xlsx", sep = ""))
}

# Visualize some genes
ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  gene = logcounts(sce)[rowData(sce)$Symbol == "PDZD3",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = gene)) + scale_colour_viridis()

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  gene = logcounts(sce)[rowData(sce)$Symbol == "BIRC5",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = gene)) + scale_colour_viridis()
```

# Monocle

I will run monocle on cell from patient 3.

```{r}
library(monocle)
cur_sce <- sce[,sce$Patient == "Patient3"]
cur_sce <- cur_sce[Matrix::rowMeans(counts(cur_sce)) > 0,]
cur_sce <- normalize(cur_sce)
HVgenes <- HVG(cur_sce)

# Calculate entropy
cur_counts <- 2^(logcounts(cur_sce)) - 1 + 0.1
probs   <- t(t(cur_counts)/apply(cur_counts,2,sum))
entropy <- -apply(probs*log(probs)/log(nrow(cur_sce)),2,sum)

# Calculate monocle
pd <- new("AnnotatedDataFrame", data = data.frame(clusters = as.factor(colData(cur_sce)$Tissue_cluster),
                                                  entropy = entropy))
fd <- new("AnnotatedDataFrame", data = data.frame(gene_short_name = rowData(cur_sce)$Symbol,
                                                  row.names = rowData(cur_sce)$ID))
cds <- newCellDataSet(counts(cur_sce), 
                      phenoData = pd, featureData = fd, expressionFamily = negbinomial.size())
cds <- estimateSizeFactors(cds)
cds <- estimateDispersions(cds)
cds <- setOrderingFilter(cds, HVgenes)
plot_ordering_genes(cds)

cds <- reduceDimension(cds,  max_components = 2, method = 'DDRTree')
cds <- orderCells(cds)

plot_cell_trajectory(cds, color_by = "clusters")
plot_cell_trajectory(cds, color_by = "entropy") + scale_color_gradientn(colours = magma(100))
plot_cell_trajectory(cds, color_by = "Pseudotime")
plot_cell_trajectory(cds, color_by = "State")

diff_test_res <- differentialGeneTest(cds,
fullModelFormulaStr = "~sm.ns(Pseudotime)")
diff_test_res <- diff_test_res[order(diff_test_res$qval, decreasing = FALSE),]
plot_genes_in_pseudotime(cds[match(diff_test_res$gene_short_name[1:5], featureData(cds)$gene_short_name),], 
                         color_by = "clusters")
```