---
title: "Trajectory analysis"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, 
        encoding = encoding, output_file = '../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/Results/Reports/Trajectory_D2.html') })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(reticulate)
use_python("/usr/bin/python")
mpl = reticulate::import("matplotlib")
mpl$use('TkAgg')
```

This script performs trajectory analysis on the BE and GC samples

## Read in batch corrected data

```{r}
suppressMessages(library(scran))
suppressMessages(library(Rtsne))
suppressMessages(library(ggplot2))
suppressMessages(library(viridis))
suppressMessages(library(monocle))
suppressMessages(library(plot3Drgl))
suppressMessages(library(umap))
suppressMessages(library(vegan))
suppressMessages(source("../Functions/auxiliary.R"))
sce <- readRDS("../../../../Dropbox (Cambridge University)/Oesophagus_single_cell/All_corrected_sce.rds")

# Select only D2
sce <- sce[,colData(sce)$Tissue == "D2"]
sce <- sce[Matrix::rowMeans(counts(sce)) > 0.1,]

# Split sce object
sce_single <- split.sce(sce, groups = unique(colData(sce)$Sample), colData.name = "Sample")

corrected <- batch.correction(sce_single)

sce <- do.call("cbind", sce_single)

colnames(corrected) <- paste(colData(sce)$Barcode, colData(sce)$Patient, colData(sce)$Tissue, sep = "_")
```

# Try different visualization techniques

## PCA

```{r}
# Batch corrected data already come as pca
ggplot(data.frame(PC1 = corrected[1,],
                  PC2 = corrected[2,],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = PC1, y = PC2, colour = patient))

ggplot(data.frame(PC1 = corrected[1,],
                  PC2 = corrected[2,],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = PC1, y = PC2, colour = cluster))
```

## tSNE

```{r}
set.seed(12345)
tsne <- Rtsne(t(corrected), pca = FALSE)

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = cluster))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = patient))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = tsne1, y = tsne2, colour = OLFM4)) + scale_colour_viridis()

```

# Diffusionmap

```{r}
library(destiny)

dm <- DiffusionMap(t(corrected), k = 20)

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = DC1, y = DC2, color = patient))

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = DC1, y = DC2, color = cluster))

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = DC1, y = DC2, colour = OLFM4)) + scale_colour_viridis()
```

# PHATE

```{r}
library(phateR)

tree.phate <- phate(t(corrected), gamma = 0)

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = phate1, y = phate2, colour = patient))

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = phate1, y = phate2, colour = cluster))

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = phate1, y = phate2, colour = OLFM4)) + scale_colour_viridis()
```

# UMAP

```{r}
umap.all <- umap(t(corrected))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  patient = colData(sce)$Patient)) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = patient))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = cluster))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = OLFM4)) + scale_colour_viridis()
```

# Entropy

The stemID score is calculated based on the cell-specific entropy times the number of links between clusters.
We can therefore calculate the cell-specific entropy as a proxy for stemness.
The entropy has to be computed on the normalized counts.

```{r entropy}
# These functions are taken from the RaceID3_StemID2_class.R script
# 1. approach - Add pseudotcount of 0.1
cur_counts <- 2^(logcounts(sce)) - 1 + 0.1
probs   <- t(t(cur_counts)/apply(cur_counts,2,sum))
entropy <- -apply(probs*log(probs)/log(nrow(sce)),2,sum)

# Cluster with highest median entropy
clust.entropy <- vector(length = length(unique(colData(sce)$Tissue_cluster)))
names(clust.entropy) <- as.character(unique(colData(sce)$Tissue_cluster))
for(i in unique(colData(sce)$Tissue_cluster)){
  clust.entropy[as.character(i)] <- mean(entropy[as.numeric(colData(sce)$Tissue_cluster) == i])
}

ggplot(data.frame(score = as.numeric(clust.entropy),
                  clust.names = names(clust.entropy))) +
  geom_point(aes(clust.names, score)) + ylim(c(0.3, 0.9))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  cluster = as.factor(colData(sce)$Tissue_cluster))) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = cluster))

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  entropy = entropy)) + 
  geom_point(aes(x = UMAP1, y = UMAP2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

ggplot(data.frame(phate1 = tree.phate$embedding[,1],
                  phate2 = tree.phate$embedding[,2],
                  entropy = entropy)) + 
  geom_point(aes(x = phate1, y = phate2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

ggplot(data.frame(DC1 = dm$DC1,
                  DC2 = dm$DC2,
                  entropy = entropy)) + 
  geom_point(aes(x = DC1, y = DC2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  entropy = entropy)) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = entropy)) +
  scale_color_gradientn(colours = magma(100))

# Correlation between stemness and OLFM4 expression
ggplot(data.frame(entropy = entropy,
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(entropy, OLFM4))

# Find marker genes for stem cells
cors <- t(apply(logcounts(sce), 1, function(n){
  cur_cor <- suppressWarnings(cor.test(n, entropy, method = "spearman"))
  c(cur_cor$estimate, cur_cor$p.value)
}))
cors <- as.data.frame(cors)
cors$BH <- p.adjust(cors$V2, method = "BH")
cors$Symbol <- rowData(sce)$Symbol[match(rownames(cors), rowData(sce)$ID)]

# Order correlations
cors.pos <- cors[cors$rho > 0 & cors$BH < 0.01,]
head(cors.pos[order(cors.pos$rho, decreasing = TRUE),], n = 100)

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  EEF1A1 = logcounts(sce)[rowData(sce)$Symbol == "EEF1A1",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = EEF1A1)) + scale_colour_viridis()

ggplot(data.frame(UMAP1 = umap.all$layout[,1],
                  UMAP2 = umap.all$layout[,2],
                  GSTA1 = logcounts(sce)[rowData(sce)$Symbol == "GSTA1",])) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = GSTA1)) + scale_colour_viridis()
```

# Monocle

```{r monocle, eval = FALSE}
cds <- monocle::newCellDataSet(cellData = corrected)
```

# StemID

```{r stem_id, eval=FALSE}
source("../../../RaceID3_StemID2/RaceID3_StemID2_class.R")

# Run StemID on cells from Patient3
sce <- sce[,sce$Patient == "Patient3"]
sce <- sce[Matrix::rowMeans(counts(sce)) > 0.1,]

# Find highly variable genes
HVgenes <- HVG(sce)

sc <- SCseq(as.data.frame(as.matrix(counts(sce)[HVgenes,])))
# For normalized data, pseudocount of 0.1 is added for entropy calculation
sc@ndata <- as.data.frame(as.matrix(2^(logcounts(sce)[HVgenes,]) - 1 + 0.1))
sc <- clustexp(sc, FSelect = FALSE)
sc <- findoutliers(sc)
sc <- rfcorrect(sc,rfseed=12345,final=TRUE,nbfactor=5)
sc <- comptsne(sc,rseed=15555,sammonmap=FALSE,initial_cmd=TRUE)

ggplot(data.frame(tsne1 = sc@tsne$V1,
                  tsne2 = sc@tsne$V2,
                  cluster = as.factor(sc@cluster$kpart))) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = cluster))

ggplot(data.frame(tsne1 = sc@tsne$V1,
                  tsne2 = sc@tsne$V2,
                  cluster = as.factor(sce$Tissue_cluster))) + 
  geom_point(aes(x = tsne1, y = tsne2, colour = cluster))

ggplot(data.frame(tsne1 = sc@tsne$V1,
                  tsne2 = sc@tsne$V2,
                  OLFM4 = logcounts(sce)[rowData(sce)$Symbol == "OLFM4",])) +
  geom_point(aes(x = tsne1, y = tsne2, colour = OLFM4)) + scale_colour_viridis()

ltr <- Ltree(sc)
ltr <- compentropy(ltr)
ltr <- projcells(ltr, cthr = 0, nmode = FALSE)
ltr <- projback(ltr, fast = TRUE)
ltr <- lineagetree(ltr,pthr=0.01, nmode=FALSE, fast=TRUE)
ltr <- comppvalue(ltr,pethr=0.01, nmode=FALSE, fast=TRUE)
plotmap(ltr)
plotlinkpv(ltr)
x <- compscore(ltr,nn=1,scthr=.3)

# Plot the StemID score for each cluster
ggplot(data.frame(score = x$StemIDscore,
                  clust.names = names(x$StemIDscore))) +
  geom_col(aes(clust.names, score))
```


